\documentclass{beamer}
\usepackage{listings}
\usepackage{amsmath}

\usetheme{metropolis}

\setmonofont[
   Numbers =,
   Scale = 0.9,
   Extension = .otf,
   WordSpace = {1,0,0},
   PunctuationSpace = WordSpace,
   UprightFont =*-Regular,
   ItalicFont =*-RegularIt,
   BoldFont=*-Bold,
   BoldItalicFont=*-BoldIt
 ]{SourceCodePro}

\lstset{
  breaklines=false,%true,
  basicstyle=\normalsize\ttfamily,
}

\title{{\small Greenlight Presentation Msc Thesis}\\Elaine: Elaborations of Higher-Order Effects as First-Class Language Feature}
\author{Terts Diepraam \\ Supervisors: Casper Bach Poulsen \& Cas van der Rest \\ Thesis advisor: Matthijs Spaan}

\begin{document}

\begin{frame}
    \titlepage
\end{frame}

\begin{frame}{Effects}
    \begin{itemize}
        \item Printing,
        \item State,
        \item Filesystem access,
        \item Coroutines,
        \item Exceptions,
        \item \& everything else monads can do.
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{Effects}
Many programming languages have limitations when dealing with effects:
\begin{itemize}
    \item the effects are not reflected in the type system,
    \item the effects are not modular.
\end{itemize}

What they do have is limited to specific features (e.g. exceptions).

\begin{lstlisting}[language=python]
def foo(x: int) -> int: # <- does not include effect
    print(x)            # <- implementation is fixed
    return x * x
\end{lstlisting}
\end{frame}

\begin{frame}{Algebraic Effects}

In languages with algebraic effects, we can:

\begin{itemize}
\item declare custom effects,
\item and define \textbf{handlers} for them.
\end{itemize}

Effects need to satisfy the \emph{algebraicity property}.

Languages: Koka, Frank, Eff, and more.

Example effects: state, IO, iteration, nondeterminism
\end{frame}

\begin{frame}{Higher-Order Effects}

Higher-order effects: effect operations that take effectful computations as parameters.

They are not algebraic: we cannot make them modular.

Example effects: Reader monad with local, exception catching, lambda abstraction

\end{frame}

\begin{frame}{Hefty Algebras}

Higher-order effects are \textbf{elaborated} into expressions with only algebraic effects.

The elaboration provides modularity we want.

\end{frame}

\begin{frame}{Elaine}
Language designed in this thesis

Core idea: apply hefty algebras to PL design

\vspace{2em}

Novel features:

\begin{itemize}
    \item elaborations as language construct
    \item implicit elaboration resolution
    \item elaboration compilation
\end{itemize}
\end{frame}

\end{document}